---
title: "Homework 3 for Network Data Analysis"
author: "Daniel P. Hall Riggins, MD"
date: "4/27/2022"
bibliography: network_citations.bib
output: html_document
---

## Prep

Load needed dependencies:

```{r message=FALSE, warning=FALSE}
library(igraph)
library(ggdist)
library(ggraph)
library(networkD3)
library(targets)
library(tidygraph)
library(tidyverse)
source("R/wrangling_generic_graphs.R")
```

Ensure data pipeline is fresh and load it:

```{r}
tar_make()
tar_load(MUC_huttlin_graph)
```

Next, I wrangle the graph for use in this assignment. If interested, see the [source code here](https://github.com/andtheWings/network_data_analysis/blob/main/R/wrangling_generic_graphs.R) for converting a graph to its maximally connected component.

```{r message=FALSE, warning=FALSE}
maximal_muc_huttlin <- 
    MUC_huttlin_graph |> 
    # As instructed, use the undirected version of the graph for this homework
    to_undirected() |> 
    # And only use the maximally connected component
    to_max_connected_component()
```

## Overall Network Topologies

Visualize the graph derived from @huttlinDualProteomescaleNetworks2021. Just for fun, I'm doing it as an interactive plot this time:

```{r}
simpleNetwork(
    as_tibble(
        activate(maximal_muc_huttlin, edges)
    ),
    Source = "official_symbol_from",
    Target = "official_symbol_to",
    zoom = TRUE
)
```


For comparison of overall topology, we will also make a random graph using the Erdos-Renyi algorithm.

```{r message=FALSE, warning=FALSE}
maximal_erdos <-
    play_erdos_renyi(
        # Use the same number of nodes and edges as in the PPI graph
        n = gorder(MUC_huttlin_graph),
        m = gsize(MUC_huttlin_graph),
        directed = FALSE
    ) |> 
    to_max_connected_component()


simpleNetwork(
    as_tibble(
        activate(maximal_erdos, edges)
    ),
    Source = "from",
    Target = "to",
    zoom = TRUE
)
```

Our `maximal_MUC_huttlin` graph looks much more organized than the `maximal_erdos` graph despite them both deriving from parent graphs with the same number of nodes and edges

### Transitivity and Mean Geodesic Distance

Generate stats about clustering and path lengths:

```{r}
transitivity_and_geodesic <-
    tibble(
        graph_name = c("MUC_huttlin", "erdos"),
        tbl_graph_obj = list(maximal_muc_huttlin, maximal_erdos)
    ) |> 
    mutate(
        transitivity = map(tbl_graph_obj, ~transitivity(.x)),
        mean_distance = map(tbl_graph_obj, ~mean_distance(.x))
    ) |> 
    unnest(c(transitivity, mean_distance))

transitivity_and_geodesic
```

`maximal_MUC_huttlin` has a higher clustering coefficient and shorter mean geodesic distance than the `erdos` graph. According to class notes, we expect that if we had a "Small World" graph, than it would indeed have a higher clustering coefficient, but we should have a lower mean path length, which we do not.

### Degree

Plot the degree distribution of `maximal_MUC_huttlin`:

```{r}
maximal_muc_huttlin |> 
    activate(nodes) |> 
    mutate(degree = centrality_degree()) |> 
    as_tibble() |> 
    group_by(degree) |> 
    summarize(degree_count = n()) |> 
    mutate(
        degree_proportion = degree_count / sum(degree_count),
        log_degree = log10(degree),
        log_degree_proportion = log10(degree_proportion)
    ) |> 
    ggplot(
        aes(
            x = log_degree,
            y = log_degree_proportion
        )
    ) +
    geom_point() 
```

The beginning portion of this log-log tranformed plot has a linear slope consistent with a scale-free graph. The visual topology of the `maximal_MUC_huttlin` graph also seems consistent with scale-free.

```{r}
maximal_erdos |> 
    activate(nodes) |> 
    mutate(degree = centrality_degree()) |> 
    as_tibble() |> 
    group_by(degree) |> 
    summarize(degree_count = n()) |> 
    mutate(
        degree_proportion = degree_count / sum(degree_count),
        log_degree = log10(degree),
        log_degree_proportion = log10(degree_proportion)
    ) |> 
    ggplot(
        aes(
            x = log_degree,
            y = log_degree_proportion
        )
    ) +
    geom_point()
```

In contrast, the log-log plot of the random graph follows a more Poisson-like distribution.

## Local network topology descriptors

### Gursoy et al. (2008)

Definitions:

- Hub protein - highly connected protein
- Closeness - how quickly information can be transferred in a network
- Betweenness centrality - nodes important to information transfer by mediating many shortest paths
- HBLC - nodes that mediate many shortest paths, but themselves are not highly connected (connectors between modules)

### Hubs

Identify the nodes with the top 10 highest eigen scores. According to the igraph manual:

> In general, vertices with high eigenvector centralities are those which are connected to many other vertices which are, in turn, connected to many others (and so on).

```{r}
top_10_hubs <-
    maximal_muc_huttlin |> 
    activate(nodes) |> 
    mutate(hub_score = centrality_eigen()) |> 
    arrange(desc(hub_score)) |> 
    as_tibble() |> 
    select(official_symbol, hub_score) |> 
    head(10)
    
top_10_hubs
```

Visualize the graph without these top 10 hub nodes:

```{r}
maximal_muc_huttlin |>
    anti_join(top_10_hubs, by = "official_symbol") |> 
    ggraph(layout='igraph', algorithm='fr') +
    geom_node_point() +
    geom_edge_link() +
    theme_graph()
```

Removing these nodes from the graph makes it more disjoint. From physiologic perspective, these are likely proteins whose roles are relevant to multiple types of cellular function important to formation of a normal umbilical cord. For example, the protein with the highest hub score is LGALS1, which is a Galectin 1, which is a important to proper scaffolding of blood vessels in angiogenesis. 

### Closeness

Identify the nodes with the top 10 highest closeness scores:

```{r}
top_10_close <-
    maximal_muc_huttlin |> 
    activate(nodes) |> 
    mutate(closeness_score = centrality_closeness()) |> 
    arrange(desc(closeness_score)) |> 
    as_tibble() |> 
    select(official_symbol, closeness_score) |> 
    head(10)

top_10_close
```

Visualize the position of these nodes in the network:

```{r}
maximal_muc_huttlin |>
    mutate(top_10_close = official_symbol %in% top_10_close$official_symbol) |> 
    ggraph(layout='igraph', algorithm='fr') +
    geom_node_point(
        aes(
            shape = top_10_close,
            size = top_10_close
        )
    ) +
    scale_shape_manual(values = c("circle", "diamond plus")) +
    scale_size_manual(values = c(1, 3)) +
    geom_edge_link(alpha = 0.3) +
    theme_graph()
```

Nodes with high closeness are important to efficient information transfer within a network. Perhaps in this context, they represent proteins that act as mediators between functional classes of proteins important to umbilical cord formation. For example, GRID1 is part of glutamate receptor for a channel protein that propagates a signal sent to a cell--perhaps starting cascades for blood vessel or smooth muscle formation.

### Betweenness

Identify the nodes with the top 10 highest betweenness scores:

```{r message=FALSE, warning=FALSE}
top_10_between <-
    maximal_muc_huttlin |> 
    activate(nodes) |> 
    mutate(betweenness_score = centrality_betweenness()) |> 
    arrange(desc(betweenness_score)) |> 
    as_tibble() |> 
    select(official_symbol, betweenness_score) |> 
    head(10)

top_10_between
```

Visualize the position of these nodes in the network:

```{r}
maximal_muc_huttlin |> 
    mutate(
        in_top_10_close = official_symbol %in% top_10_close$official_symbol,
        in_top_10_between = official_symbol %in% top_10_between$official_symbol,
        of_interest = official_symbol %in% c(top_10_close$official_symbol, top_10_between$official_symbol)
    ) |> 
    ggraph(layout='igraph', algorithm='fr') +
    geom_node_point(
        aes(
            shape = in_top_10_close,
            color = in_top_10_between,
            size = of_interest
        )
    ) +
    scale_shape_manual(values = c("circle", "diamond plus")) +
    scale_size_manual(values = c(1, 3)) +
    scale_color_manual(values = c("black", "red")) +
    geom_edge_link(alpha = 0.2) +
    theme_graph()
```

We see that 5 of the nodes are the top 10 for both closeness (red color) and betweenness (diamond shape). This is not surprising as they serve similar roles, but I liked this way of differentiating them [from Visible Network Labs](https://visiblenetworklabs.com/2021/04/16/understanding-network-centrality/):

> Centrality is a helpful measure for identifying key players in a network. Depending on the specific measure used, centrality means a network is directly connected to many others (degree centrality), close to many others indirectly (closeness centrality), or serve as a key broker between many other nodes (betweenness centrality).

I interpret this to mean that the broker provides connections between clusters that otherwise would not be as closely related. For example, BLVRB is an enzyme in the liver that helps in the catabolism of disparate molecules like biliverdin and riboflavin. Perhaps in this context, it suggests that malformed uterine cords lead to higher demand for waste products to be broken down.

### HBLCs

Here is a rough means of identifying HBLC's by finding those that are in the top 10 for betweenness, but not for closeness or hub score:

```{r message=FALSE, warning=FALSE}
hblcs <-
    maximal_muc_huttlin |>
    mutate(
        hub_score = centrality_eigen(),
        closeness_score = centrality_closeness(),
        betweenness_score = centrality_betweenness()
    ) |>
    as_tibble() |> 
    select(
        official_symbol,
        hub_score,
        closeness_score,
        betweenness_score
    ) |> 
    semi_join(top_10_between) |> 
    anti_join(top_10_close) |> 
    anti_join(top_10_hubs) |> 
    arrange(
        desc(betweenness_score),
        hub_score,
        closeness_score
    )

hblcs
```

SERPINB5 is the most likely candidate by this method. Ironically, [scientific consensus](https://en.wikipedia.org/wiki/Maspin) cannot agree on a specific role for this protein, although it may be involved in tumor suppression.

Here are the candidates on the graph (triangles):

```{r}
maximal_muc_huttlin |> 
    mutate(
        shape_var =
            case_when(
                official_symbol %in% top_10_close$official_symbol ~ "close",
                official_symbol %in% hblcs$official_symbol ~ "hblc",
                TRUE ~ "neither"
            ),
        in_top_10_between = official_symbol %in% top_10_between$official_symbol,
        of_interest = 
            official_symbol %in% c(
                top_10_close$official_symbol, 
                top_10_between$official_symbol,
                hblcs$official_symbol
            )
    ) |> 
    ggraph(layout='igraph', algorithm='fr') +
    geom_node_point(
        aes(
            shape = shape_var,
            color = in_top_10_between,
            size = of_interest
        )
    ) +
    scale_shape_manual(values = c("diamond plus", "triangle", "circle")) +
    scale_size_manual(values = c(1, 3)) +
    scale_color_manual(values = c("black", "red")) +
    geom_edge_link(alpha = 0.2) +
    theme_graph()
```

Looking at the HBLCs, they seem to be acting as brokers to clusters of nodes that would likely otherwise not be on the graph. Physiologically, these HBLC's might be top candidates for proteins essential for development of the malformed umbilical cord phenotype (or for its prevention).

# Citations {-}
