---
title: "Homework 5 for Network Data Analysis"
author: "Daniel P. Hall Riggins, MD"
date: "`r Sys.Date()`"
bibliography: network_citations.bib
output: html_document
---

## Prep

```{r message=FALSE, warning=FALSE}
library(corpcor)
library(dplyr)
library(ggraph)
library(huge)
library(igraph)
library(readr)
library(targets)
library(tidygraph)
library(tidyr)
source("R/describing_generic_data.R")
```

Ensure underlying data pipeline is fresh:

```{r}
tar_make()
```

## The Dataset

I derived my dataset from the study entitled, "[Profiling Plasmodium falciparum parasites and human red blood cells after treatment with MMV693183](https://www.metabolomicsworkbench.org/data/DRCCMetadata.php?Mode=Study&DataMode=AllData&StudyID=ST001985&StudyType=MS&ResultType=1#DataTabs)" by Llinas et al. (2021). The authors set out to examine metabolomic changes in human red blood cells infected with malaria when they were treated with a novel anti-malarial compound. 

Samples were differentiated by the following variables:

- Date of assay
- Whether the sample contained the treatment compound (MMV693183)
- If it contained the compound, at what dosage (2.4 [IC50], 24, or 240 nM)
- The sample target (None, Uninfected RBCs, Infected RBCs)

Their analysis platform was untargeted LC/ESI-MS using the Thermo Orbitrap.

There are 196 metabolites whose abundance are measured.

## Data Prep

Per usual, I incorporated the data prep stages into my {[targets](https://docs.ropensci.org/targets/)} [pipeline](https://github.com/andtheWings/network_data_analysis/blob/main/_targets.R). Custom functions can be found in [my Github repo](https://github.com/andtheWings/network_data_analysis/blob/main/R/wrangling_llinas.R).

First, I transposed the data from containing metabolites in rows and sample ID's in columns:

```{r}
llinas_raw <- read_csv("data/llinas_2021_metabolomics.csv")

llinas_raw
```

To the opposite:

```{r eval=FALSE}
transposed_llinas_raw <-
    llinas_raw |> 
    tidyr::pivot_longer(
        cols = 2:73,
        names_to = "id"
    ) |> 
    tidyr::pivot_wider(
        names_from = Samples,
        values_from = value
    )
```

```{r include=FALSE}
tar_load(transposed_llinas_treatments_raw)
```

```{r}
transposed_llinas_treatments_raw
```

Then I exported the data and manually added columns to encode different pieces of information in the id column:

```{r eval=FALSE}
write_csv(transposed_llinas_raw, "data/transposed_llinas_2021.csv")

llinas_treatments_modified <- read_csv("data/expanded_transposed_llinas_2021.csv")
```

```{r}
box::use(riggins/graph_generics/wrangling_generic_graphs)
```

```{r include=FALSE}
tar_load(llinas_treatments_modified)
```

```{r}
llinas_treatments_modified
```
 
Next, I identified which metabolite variables had over 10% of data missing using a custom function. There were 2:

```{r}
missing_vars <- find_vars_w_missing_data_above_threshold(llinas_treatments_modified, 0.1)

missing_vars
```

Next, I prepared my metabolite matrix by removing all non-metabolite columns/the metabolite columns with missing data and standardizing the abundance measurements:

```{r eval=FALSE}
llinas_treatments_pre_ggm <-
        llinas_treatments_modified |> 
        select(
            -id,
            -date,
            -treatment,
            -target,
            -dosage,
            -replicate,
            -missing_vars
        ) |> 
        datawizard::standardise() |> 
        as.matrix()
```

```{r include=FALSE}
tar_load(llinas_treatments_pre_ggm)
```

```{r}
llinas_treatments_pre_ggm[1:5,1:5]
```

One final note: I also manually added a variable to the table of metabolite nodes that assigned a broad biomolecular category to each metabolite using the [ChEBI Ontology](https://bioportal.bioontology.org/ontologies/CHEBI).

```{r}
tar_load(llinas_metabolite_nodes)

llinas_metabolite_nodes
```

## Graphical Guassian Model (GGM) Estimation

I inferred edge relationships between the metabolites using the {[huge](https://cran.r-project.org/web/packages/huge/)} package's graphical lasso method:

```{r eval=FALSE}
llinas_treatments_ggm <- huge(llinas_treatments_pre_ggm, method = "glasso", cov.output=TRUE)
```

```{r include=FALSE}
tar_load(llinas_treatments_ggm)
```

```{r}
llinas_treatments_ggm
```

I used two methods for selecting the threshold parameter's value. Rotation information criterion (RIC):

```{r eval=FALSE}
llinas_treatments_ggm_ric <- huge.select(llinas_treatments_ggm, criterion="ric", cov.output=TRUE)
```

```{r include=FALSE}
tar_load(llinas_treatments_ggm_ric)
```

```{r}
llinas_treatments_ggm_ric
```
Which yielded a sparsity level of 0.24:

```{r}
plot(llinas_treatments_ggm_ric)
```

```{r}
llinas_treatments_ggm_ric$refit |> 
        as_tbl_graph() |> 
        activate(edges) |> 
        as_tibble()
```


```{r}
llinas_metabolite_ric_pcors1 <-
        llinas_treatments_ggm_ric_obj$opt.cov |> 
        cov2cor() |> 
        corpcor::cor2pcor() |> 
        as_tbl_graph() |> 
        activate(edges) |> 
        as_tibble() |> 
        semi_join(
            llinas_metabolite_ric_edges1,
            by = c("from", "to")
        )
```


And the stability approach for regularization selection (StARs):

```{r eval=FALSE}
llinas_treatments_ggm_stars <- huge.select(llinas_treatments_ggm, criterion="stars")
```

```{r include=FALSE}
tar_load(llinas_treatments_ggm_stars)
```

```{r}
llinas_treatments_ggm_stars
```

Which was unsuccessful and yielded a sparsity level of 0.

```{r}
plot(llinas_treatments_ggm_stars)
```

Next I assembled my final graph using a custom function:

```{r eval=FALSE}
llinas_metabolite_graph <- assemble_llinas_metabolite_graph(llinas_metabolite_nodes, llinas_treatments_ggm_ric)
```

```{r include=FALSE}
tar_load(llinas_metabolite_graph)
```

And graphed the inferred network, coloring nodes by the broad biomolecular type, and weighting edges by their partial correlation values:

```{r message=FALSE, warning=FALSE}
ggraph(
        llinas_metabolite_graph
    ) +
    geom_edge_link(
        aes(
            color = abs(weight)
        ),
        alpha = 0.6
    ) +
    scale_edge_color_gradient(
        name = "Partial Correlation",
        low = "white",
        high = "black"
    ) +
    geom_node_point(
        aes(
            color = chem_cat
        )
    ) +
    labs(
        title = "Inferred metabolite relationships in the \n Llinas experimental assays",
        color = "Molecule Type"
    ) +
    theme_graph()
```

