---
title: "Homework 1 for Network Data Analysis"
author: "Daniel P. Hall Riggins, MD"
date: "4/12/2022"
output: html_document
---

Load needed libraries:

```{r include=FALSE}
library(dplyr)
library(tidygraph)
library(ggraph)
library(stringr)
library(tidyr)
```

## Make the Graph

Read in BioGRID's human-specific interaction dataset:

```{r}
raw <- readr::read_delim("data/BIOGRID-ORGANISM-Homo_sapiens-4.4.207.tab3.txt")
```

Show datasets with the most interactions:

```{r}
janitor::tabyl(raw, `Publication Source`) |> 
    arrange(desc(n)) |> 
    head(10)
```

The publication source "PUBMED:33961781" corresponds to [Huttlin et al. 2021](https://pubmed.ncbi.nlm.nih.gov/33961781/). This research publication contains information on 14,586 unique proteins and 118,162 interactions for human cell lines 293T (embryonic kidney origin) and HCT116 (colon cancer origin). The researchers detected interactions using affinity-purification mass spectrometry (AP-MS), which is a selective method for characterizing macromolecular binding, resulting in a low false discovery rate, but likely omitting many weak or transient protein interactions.

Build a graph of the Huttlin et al. (2021) data using the {Tidygraph} API for igraph:

```{r}
# Filter for only the Huttlin data source
huttlin <- 
    raw |> 
    filter(`Publication Source` %in% "PUBMED:33961781") |> 
    separate(
        col = Qualifications,
        into = c(NA, NA, "cell_line"),
        sep = " ",
        remove = FALSE
    ) |> 
    mutate(
        cell_line = 
            case_when(
                cell_line %in% "HEK" ~ "293T",
                cell_line %in% "HCT116" ~ "HCT116"
            )
    )

# Assemble a nodes dataframe from the "A" interactors
huttlin_nodes_A <-
    huttlin |> 
    select(
        cell_line,
        official_symbol = `Official Symbol Interactor A`,
        entrez = `Entrez Gene Interactor A`,
        biogrid_interactor = `BioGRID ID Interactor A`,
        systematic_name = `Systematic Name Interactor A`,
        synonyms = `Synonyms Interactor A`,
        swiss_prot = `SWISS-PROT Accessions Interactor A`,
        trembl = `TREMBL Accessions Interactor A`,
        refseq = `REFSEQ Accessions Interactor A`
    ) |> 
    # `Entrez Gene Interactor A` was parsed incorrectly as character type
    mutate(
        entrez = as.numeric(entrez)
    ) |> 
    unite(
        col = cell_line_and_official_symbol,
        cell_line, official_symbol,
        remove = FALSE
    )
    

# Assemble a nodes dataframe from the "B" interactors then bind to the "A" interactors
huttlin_nodes <-
    huttlin |> 
    select(
        cell_line,
        official_symbol = `Official Symbol Interactor B`,
        entrez = `Entrez Gene Interactor B`,
        biogrid_interactor = `BioGRID ID Interactor B`,
        systematic_name = `Systematic Name Interactor B`,
        synonyms = `Synonyms Interactor B`,
        swiss_prot = `SWISS-PROT Accessions Interactor B`,
        trembl = `TREMBL Accessions Interactor B`,
        refseq = `REFSEQ Accessions Interactor B`
    ) |> 
    unite(
        col = cell_line_and_official_symbol,
        cell_line, official_symbol,
        remove = FALSE
    ) |> 
    bind_rows(huttlin_nodes_A) |> 
    # Remove redundant observations
    distinct()



# Assemble an edges dataframe
huttlin_edges <-
    huttlin |> 
    unite(
        col = "from",
        cell_line, `Official Symbol Interactor A`,
        remove = FALSE
    ) |>
    unite(
        col = "to",
        cell_line, `Official Symbol Interactor B`,
        remove = FALSE
    ) |> 
    select(
        from,
        to,
        cell_line,
        official_symbol_from = `Official Symbol Interactor A`,
        official_symbol_to = `Official Symbol Interactor B`,
        biogrid_interaction = `#BioGRID Interaction ID`,
        score = Score,
        modification = Modification,
        qualifications = Qualifications,
        tags = Tags,
        ontology_term_ids = `Ontology Term IDs`,
        ontology_term_names = `Ontology Term Names`,
        ontology_term_qualifier_ids = `Ontology Term Qualifier IDs`,
        ontology_term_qualifier_names = `Ontology Term Qualifier Names`,
        ontology_term_types = `Ontology Term Types`
    ) |> 
    mutate(score = scales::rescale(-1*as.numeric(score), to = c(1,10)))
           


# Assemble the graph object
huttlin_graph <- 
    tbl_graph(
        nodes = huttlin_nodes, 
        edges = huttlin_edges
    )


huttlin_graph
```

The number of nodes and edges in the graph object reflect similar orders of magnitude to those reported in the publication:

| source       | nodes | edges  |
|--------------|-------|--------|
| publication  | 14586 | 118162 |
| graph object | 14511 | 169808 |

Sidenote: The graph object retains class properties both from the tidygraph API (tbl_graph) **and** the igraph backend (igraph):

```{r}
class(huttlin_graph)
```
## Plot the graph

This graph is so large, my laptop crashes when trying to plot the whole thing. Instead we'll plot the local neighborhood subgraph of the most abundant bait protein.

List most common bait proteins:

```{r}
janitor::tabyl(huttlin_edges, from) |> 
    arrange(desc(n)) |> 
    head(10)
```
Extract the local neighborhood of the GPR182 protein:

```{r}
GPR182_graph <-
    huttlin_graph |> 
    convert(
        to_local_neighborhood,
        node = which(.N()$official_symbol == "GPR182"),
        order = 1,
        mode = "all"
    )

```

Create a simple plot of the subgraph:

```{r}
ggraph(GPR182_graph, layout='igraph', algorithm='nicely') + 
    geom_edge_link() + 
    geom_node_point(aes(color = cell_line))
```

```{r}
saveRDS(huttlin_graph, "data/huttlin_graph.RDS")
```

